Forcing for Isabelle

### 2018-12-29 LICS/conclusion.tex
parrafos sueltos

We first discuss the axioms we are now missing. The proof of the
Powerset Axiom does not present difficulties, and it is one of the
first goals in the immediate future. This is also the case with the
Axiom of Infinity; it can be proved in a direct but cumbersome way. We
preferred to wait to have a full-fledged inteface between ctms and the
locales providing recursive constructions from Paulson's
\isatt{Constructibility} session. Then the current proof will hold
with no extra assumptions. The same goes for the results $M\sbq M[G]$
and $G\in M[G]$.

The Replacement Axiom, however, requires some more work to be
done. Both in Kunen and in other presentations \cite{neeman-course}
it requires a relativized version (i.e., showing that it holds for $M$) of
the \emph{Reflection Principle}. In order to state this meta-theoretic
result by Montague, recall that
an equivalent formulation of the Foundation Axiom states that the 
universe of sets can be decomposed in a transfinite hierarchy of
sets. 
\begin{theorem}
  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
  $\al$. Then each $V_\al$ is a set and 
  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
\end{theorem}
\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
  unboundedly many $\al$ such that $V_\al\models \Phi$.''
\end{theorem}

This result has also other applications; in particular, it dispenses
the need of ctms of the whole of $\ZF$ for consistency proofs (by
repeating  the proof of Lemma~\ref{lem:wf-model-implies-ctm}, now
starting with a  $V_\al$ satisfying $\Phi$).

\medskip
\fbox{Decir que Separaci\'on con 6 variables en $M$}

\fbox{implica Separaci\'on con a lo sumo 2 en $M[G]$}
\medskip

>>>>

We first discuss the axioms we are now missing. The proof of the
Powerset Axiom does not present new difficulties but  has to
take care of absoluteness issues; it is one of the
first goals in the immediate future. This is also the case with the
Axiom of Infinity; it can be proved in a direct but cumbersome way. We
preferred to wait to have a full-fledged inteface between ctms and the
locales providing recursive constructions from Paulson's
\isatt{Constructibility} session. Then the current proof will hold
with no extra assumptions. The same goes for the results $M\sbq M[G]$
and $G\in M[G]$.

The Replacement Axiom, however, requires some more work to be
done. Both in Kunen and in other presentations \cite{neeman-course}
it requires a relativized version (i.e., showing that it holds for $M$) of
the \emph{Reflection Principle}. In order to state this meta-theoretic
result by Montague, recall that
an equivalent formulation of the Foundation Axiom states that the 
universe of sets can be decomposed in a transfinite hierarchy of
sets. 
\begin{theorem}
  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
  $\al$. Then each $V_\al$ is a set and 
  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
\end{theorem}
\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
  unboundedly many $\al$ such that $V_\al\models \Phi$.''
\end{theorem}
%% This result has also other applications; in particular, it dispenses
%% the need of ctms of the whole of $\ZF$ for consistency proofs (by
%% repeating  the proof of Lemma~\ref{lem:wf-model-implies-ctm}, now
%% starting with a  $V_\al$ satisfying $\Phi$).

This statement is straightforwardly equivalent to that with $\Phi$
consisting of a single formula. The schematic nature of this result
hints at a proof by induction on formulas, and hence it must be shown
internally.

This  is an appropriate spot to insist that the internal/external
dichotomy has been a powerful agent in the shaping of our
project. It traces back to the definition of the formula-transformer
$\forceisa$, which must be done by recursion, and this requirement
spills indirectly to the proof of the Separation Axiom (despite this
is not by induction). Now the need of using internalized formulas
reappears with the need of having a general version of the Reflection
Principle (because the $\Phi$ there depends on the instance of Replacement
being proved).

A secondary goal of this project is to assess which assumptions on the
original ctm $M$ are needed to develop the forcing machinery. Up to this
point we have some anecdotal data; for example, to show that a
2-variable instance of Separation holds in $M[G]$, one needs to use a
6-variable instance in $M$.
